Applying Radial Basis Function Networks and Markov Chains for on-line detection of concept drift in non-stationary environments

### 2019-10-27 tex/article/article.tex
abstract!

\title{Applying Radial Basis Networks and Markov Chains for on-line detection of concept drift in non-stationary environments}

@@

\begin{abstract}
The amount of information produced by computer systems has grown sharply in recent decades.
Most of this information is presented in the form of uninterrupted and potentially infinite sequences known as data streams.
These streams are usually produced in non-stationary environments,
in which the data distribution can change over time,
possibly deteriorating the system performance.
In the literature,
this phenomenon is named concept drift.
%
Nevertheless,
most drift detection methods are unsuited for non-stationary environments with data streams.
These algorithms or require the correct labeling of data - infeasible in these settings,
or do not match the strict response time and resource usage restrictions inherent to data streams.
%
In an attempt to mitigate the aforementioned problem,
this paper proposes a novel proactive method for on-line drift detection,
called RBFChain.
The proposed method relies on Radial Basis Networks implicit clustering property
and uses Markov Chains to model the drifts transitions.
%
To assess the proposed method as a viable concept drift detector,
an analysis of sensitivity, accuracy,
and noise tolerance was performed using synthetic datasets,
and results were compared to the most established algorithms in the literature.
%
Furthermore, the algorithm was applied to the real-world problem of eye-tracking.
A problem with impact in different areas of knowledge,
since many behavioral experiments use eye-tracking information as a relevant analysis factor.
Performed experiments reveal that RBFChain can classify fixations and saccades in real-time,
with high accuracy, noise tolerance, and using limited resources.
\end{abstract}

@@

%% main text
\section{Introduction}
\label{sec:intro}

In recent years,
the volume of data produced by computer systems has grown dramatically.
Technological advances favored this growth,
such as the pervasiveness of mobile devices,
the popularization of social networks and
the expansion of the internet of things \cite{Cohen:BigData:2009:MSN:1687553.1687576}.

A significant portion of this information is produced in the form of uninterrupted and potentially infinite sequences \cite{Aggarwal:2006:DSM:1196418}.
In literature,
sequences with these characteristics are called data streams and
are present in various fields of application such as
financial market monitoring  \cite{ZHOU:2015},
road traffic monitoring \cite{Wang:2015:EOV:2843092.2843464},
telecom network management \cite{delattre2015method},
real-time sentiment analysis  \cite{KRANJC2015187}
and intruder prevention and identification systems \cite{KENKRE:PAI:COLACO:2015}.

@@

Still, most drift detection methods are improper for non-stationary environments with data streams.
These methods or require the correct labeling of data - impracticable in these contexts,
or do not meet the severe response time and resource usage restrictions intrinsic to data streams.

This paper proposes a novel proactive method for on-line drift detection,
called RBFChain.
The proposed algorithm is based on Radial Basis Networks implicit clustering property
and employs Markov Chains to model the drifts transitions.
To validate the proposed method as a viable concept drift detector,
an examination of sensitivity, accuracy,
and noise tolerance was performed using synthetic datasets,
and results were compared to the most established algorithms in the literature.

@@

The rest of the paper is organized as follows:
Section 2 describes the concept drift phenomenon and the main detection techniques;
Section 3 presents the eye-tracking problem, with a focus on the identification of fixations and saccades;
Section 4 describes the RBFChain algorithm and its pseudo-code;
Section 5 shows the experiment configuration for synthetic datasets and the obtained results;
Section 6 presents the examination setup with a real-world eye-tracking problem and observed outcomes; and, finally, Section 7 provides conclusions
and discusses future work.

\section{Concept Drift}
\label{sec:concept_drift}

>>>>

\title{Applying Radial Basis Function Networks and Markov Chains for on-line detection of concept drift in non-stationary environments}

@@

\begin{abstract}
The amount of information produced by computer systems has grown sharply in recent decades.
Most of this information is presented in the form of uninterrupted and potentially infinite sequences known as data streams.
These streams are usually produced in non-stationary environments,
in which the data distribution can change over time,
possibly deteriorating the system performance.
In the literature,
this phenomenon is named concept drift.
%
Nevertheless,
most drift detection methods are unsuited for non-stationary environments with data streams.
These algorithms or require the correct labeling of data - infeasible in these settings,
or do not match the strict response time and resource usage restrictions inherent to scenarios with data streams.
%
In an attempt to mitigate the aforementioned problem,
this paper proposes a novel proactive method for on-line drift detection,
called RBFChain.
The proposed method relies on Radial Basis Function Networks implicit clustering property
and uses Markov Chains to model the drifts transitions.
%
To assess the proposed method as a viable concept drift detector,
an analysis of sensitivity, accuracy,
and noise tolerance was performed using synthetic datasets,
and results were compared to the most established algorithms in the literature.
%
Furthermore, the algorithm was applied to the real-world problem of eye-tracking.
A problem with impact in different areas of knowledge,
since many behavioral experiments use eye-tracking information as a relevant analysis factor.
%
Experimental results suggest that RBFChain is statistically better or equivalent to other detectors as it offers greater or equal overall classification accuracy in most situations.
Also, the technique is applicable to eye-tracking problems as it was able to identify fixations and saccades in real-time with precision comparable to state of the art.

\end{abstract}

@@

%% main text
\section{Introduction}
\label{sec:intro}
In recent years,
the volume of data produced by computer systems has grown dramatically.
Technological advances favored this growth,
such as the pervasiveness of mobile devices,
the popularization of social networks, and
the expansion of the internet of things \cite{Cohen:BigData:2009:MSN:1687553.1687576}.

A significant portion of this information is produced in the form of uninterrupted and potentially infinite sequences \cite{Aggarwal:2006:DSM:1196418}.
In literature,
sequences with these characteristics are called data streams.
These streams are present in various fields of application,
such as financial market monitoring  \cite{ZHOU:2015},
road traffic monitoring \cite{Wang:2015:EOV:2843092.2843464},
telecom network management \cite{delattre2015method},
real-time sentiment analysis  \cite{KRANJC2015187}
and intruder prevention and identification systems \cite{KENKRE:PAI:COLACO:2015}.

@@

Still, most drift detection methods are unsuitable for non-stationary environments with data streams.
These methods or require the correct labeling of data - impracticable in these contexts,
or do not meet the severe response time and resource usage restrictions inherent to contexts with data streams.

This paper proposes a novel proactive method for on-line drift detection,
called RBFChain.
The proposed algorithm is based on Radial Basis Function Networks implicit clustering property
and employs Markov Chains to model the drifts transitions.
To validate the proposed method as a viable concept drift detector,
an examination of sensitivity, accuracy,
and noise tolerance was performed using synthetic datasets,
and results were compared to the most established algorithms in the literature.

@@

The rest of the paper is organized as follows:
Section 2 describes the concept drift phenomenon and the main detection techniques;
Section 3 presents the eye-tracking problem, with a focus on the identification of fixations and saccades;
Section 4 describes the RBFChain algorithm and its pseudo-code;
Section 5 shows the configuration and results for the experiment with synthetic datasets;
Section 6 presents the configuration and results for the experiment with the eye-tracking problem; and, finally, Section 7 provides conclusions and discusses future work.
\section{Concept Drift}
\label{sec:concept_drift}
